Scattering from compact objects: \\ Regge poles and the Complex Angular Momentum method

### 2019-12-12 rainbowCAM.tex
Further work on the Intro.

The time-independent scattering of planar waves in the gravitational field of a compact body has been studied in some detail since the 1960s \cite{Hildreth1964PhDT64, Matzner:1968, Vishveshwara:1970}. A substantial literature has developed on \emph{black hole scattering}, focussing on the canonical scenario of a planar wave of circular frequency $\omega$ and spin $s$ \cite{Chrzanowski:1976jb} impinging upon a black hole of mass $M$ in vacuum \cite{Hildreth1964PhDT64, Matzner:1968, Vishveshwara:1970, Mashhoon:1973zz,Fabbri:1975,Sanchez:1977vz,MatznerRyan1978,Handler:1980un,Matzner:1985rjn,Futterman:1988ni,Andersson:1995vi,Glampedakis:2001cx,Dolan:2006vj,Dolan:2007ut,Dolan:2008kf,Crispino:2009xt,Cotaescu:2014jca,Gussmann:2016mkp}. The dimensionless parameter
$
M \omega = \pi r_g / \lambda
$
encapsulates the ratio of the gravitational radius $r_g = 2GM/c^2$ to the wavelength $\lambda$ (herein we adopt geometric units such that $G=c=1$). The long-wavelength ($M \omega \ll 1$), short-wavelength ($M \omega \gg 1$) and intermediate regimes have been studied with a combination of perturbative \cite{DeLogi:1977dp,Dolan:2007ut,Guadagnini:2008ha,Sorge:2015yoa}, semi-classical \cite{Matzner:1985rjn, Anninos:1992ih} and numerical methods. The $s = 0$ (scalar) \cite{Matzner:1968,Sanchez:1977vz,Andersson:1995vi,Glampedakis:2001cx,Leite:2019eis}, $s=1/2$ (fermion) \cite{Dolan:2006vj,Cotaescu:2014jca}, $s=1$ (electromagnetic) \cite{Fabbri:1975, Crispino:2009xt, Crispino:2015gua} and $s=2$ (gravitational) cases \cite{MatznerRyan1978,Handler:1980un,Dolan:2008kf} have all been addressed.

Time-independent scattering by a compact body of radius $R$ with a regular centre, such as a neutron star or white dwarf, has received comparatively less attention. In such a scenario, an electromagnetic wave will not penetrate far inside the compact body; but on the other hand, a gravitational wave will pass through the body without much impediment from the matter distribution. Similarly, a neutron star is expected to be substantially transparent to neutrinos. Even in cases where the coupling to matter is negligible, the incident wave will nevertheless be scattered by the influence of the spacetime curvature. Thus, the resulting scattering pattern depends on $M\omega$, on the internal structure of the body, and on its inverse compactness or \emph{tenuity} $R/M$. Typical values of the tenuity are: $R/M \sim 6$ for neutron stars, $\sim 1.4-9.4 \times 10^3$ for white dwarfs, $4.7 \times 10^5$ for the Sun, and $1.4 \times 10^9$ for Earth. More speculatively, we shall also consider the case of (hypothetical) Ultra Compact Objects (UCO), whose tenuity is bounded from below by the Buchdahl bound of $R / M = 9/4$.

@@

CAM theory, originally developed for high-energy physics applications, has been successfully applied to black hole scattering scenarios. In 1994, Andersson and Thylwe \cite{Andersson:1994rk} described the scattering of scalar waves by a Schwarzschild BH by utilising CAM techniques, and they showed that, in this case, Regge poles are associated with ``surface waves'' localised near the unstable light ring at $r=3M$ (later it was shown that the corresponding quasinormal modes are Breit-Wigner type resonances generated by these surface waves \cite{Decanini:2002ha}). Recently, Folacci and Ould Elhadj have shown that CAM methods can be used for precise numerical calculations of scattering cross sections of scalar fields, electromagnetic fields \cite{Folacci:2019cmc} and gravitational waves  \cite{Folacci:2019vtt} on the Schwarzschild spacetime. Moreover, they have obtained an approximate  formula (see Sec.~IV in Ref.~\cite{Folacci:2019vtt}) that accurately describes both the `orbiting' and `glory' oscillations in black hole scattering cross sections in the short-wavelength regime. This approximation arises from applying WKB theory to characterise the Regge pole with the smallest imaginary part, and its residue. \sam{Mohamed, please correct this if wrong.}

CAM techniques have not yet been applied to compact-body scattering, it appears, with one notable exception: in 1991, Chandrasekhar and Ferrari \cite{ChandrasekharIV:1992ey} examined the Regge pole spectra for relativistic stellar models. 
Their study focussed on resonant modes: poles with small imaginary parts corresponding to trapped $w$-modes or fluid modes, and it employed Complex Angular Momentum (CAM) theory to calculate the flow of gravitational energy through a star.

 It seems reasonable to expect that CAM may also aid in our understanding of other compact objects such as UCOs and neutron stars. With this in mind, we now turn our attention to calculating the Regge pole spectrum for compact bodies.

It was recently shown \cite{Dolan:2017rtj, Stratton:2019deq} that time-independent scattering by compact bodies generates (in principle) the \emph{rainbow scattering} phenomenon in the short wavelength regime ($M \omega \gg 1$). The rainbow scattering phenomenon is linked to the formation of a cusp caustic in the incident wavefront (see Fig.~1 in Ref.~\cite{Stratton:2019deq}), and to the stationary point in the geodesic deflection function associated with a ray (that is, a null geodesic) that passes somewhat inside the body. For dilute bodies, such as stars and white dwarfs, the cusp caustic forms at a distance $d \sim R^2 / (4M)$ downstream of the body, and the rainbow angle is $\theta_r \approx 4M/R$ (e.g.~for the Sun, $d \approx 550 \, \text{a.u.}$ and $\theta_r \approx 1.8 \, \text{arcsec}$). For compact bodies such as neutron stars, however, the cusp caustic forms near the surface of the body, and the rainbow angle is sizable ($\theta_r \gtrsim 59.6^\circ$ for $R/M = 6$; see Table I in Ref.~\cite{Stratton:2019deq}). In this case, the rainbow angle is -- in principle at least -- a diagnostic of the matter distribution within the body and thus its nuclear equation of state.

@@

 \cite{Folacci:2018sef} \cite{Decanini:2011xi, Decanini:2011xw}

to study time-independent scattering by a compact body.

More literature exists on Regge pole theory applied to black hole scattering, for example They . Later,  We see here that CAM theory is a powerful approach to BH physics, that provides a clear and robust physical picture.

Now describe the application of CAM methods to black holes, and its successes ... The aim of this work is to extend the methods to compact bodies.

Describe QNMs - resonances in the time domain: \cite{Detweiler:1985zz, Kokkotas:1986gd, Chandrasekhar449, Kokkotas:1992ka, Leins:1993zz, Andersson:1995ez, Andersson1996} Review of QNMs: \cite{Kokkotas:1999bd}. Now describe how these are connected to Regge poles.

Finally a paragraph on the connections to Mie scattering by a transparent sphere.

\cite{Leite:2017zyb,Nambu:2019sqn, Folacci:2019vtt,}.

>>>>

The time-independent scattering of planar waves in the gravitational field of a compact body has been studied in some detail since the 1960s \cite{Hildreth1964PhDT64, Matzner:1968, Vishveshwara:1970}. A substantial literature has accumulated on \emph{black hole scattering}, focussing on the canonical scenario of a planar wave of circular frequency $\omega$ and spin $s$ \cite{Chrzanowski:1976jb} impinging upon a black hole of mass $M$ in vacuum \cite{Hildreth1964PhDT64, Matzner:1968, Vishveshwara:1970, Mashhoon:1973zz,Fabbri:1975,Sanchez:1977vz,MatznerRyan1978,Handler:1980un,Matzner:1985rjn,Futterman:1988ni,Andersson:1995vi,Glampedakis:2001cx,Dolan:2006vj,Dolan:2007ut,Dolan:2008kf,Crispino:2009xt,Cotaescu:2014jca,Gussmann:2016mkp}. A dimensionless parameter
$
M \omega = \pi r_g / \lambda
$
encapsulates the ratio of the gravitational radius $r_g = 2GM/c^2$ to the wavelength $\lambda$; herein we adopt geometric units such that $G=c=1$. The long-wavelength ($M \omega \ll 1$), short-wavelength ($M \omega \gg 1$) and intermediate regimes have been studied with a combination of perturbative \cite{DeLogi:1977dp,Dolan:2007ut,Guadagnini:2008ha,Sorge:2015yoa}, semi-classical \cite{Matzner:1985rjn, Anninos:1992ih} and numerical methods. The $s = 0$ (scalar) \cite{Matzner:1968,Sanchez:1977vz,Andersson:1995vi,Glampedakis:2001cx,Leite:2019eis}, $s=1/2$ (fermion) \cite{Dolan:2006vj,Cotaescu:2014jca}, $s=1$ (electromagnetic) \cite{Fabbri:1975, Crispino:2009xt, Crispino:2015gua} and $s=2$ (gravitational) cases \cite{MatznerRyan1978,Handler:1980un,Dolan:2008kf} have all been addressed.

Time-independent scattering by a compact body of radius $R$ with a regular centre, such as a neutron star or white dwarf, has received less attention, by comparison. In such a scenario, an electromagnetic wave will not penetrate far inside the compact body; but on the other hand, a gravitational wave will pass through the body without much impediment from the matter distribution. Similarly, a neutron star is expected to be substantially transparent to neutrinos. Even in cases where the coupling to matter is negligible, the incident wave will nevertheless be scattered by the influence of the spacetime curvature. Thus, the resulting scattering pattern depends not just on $M\omega$, but also on the internal structure of the body, and on its inverse compactness or \emph{tenuity}, $R/M$. Typical values of the tenuity are: $R/M \sim 6$ for neutron stars, $\sim 1.4-9.4 \times 10^3$ for white dwarfs, $4.7 \times 10^5$ for the Sun, and $1.4 \times 10^9$ for Earth. More speculatively, we shall also consider here the case of (hypothetical) \emph{ultra-compact objects} (UCOs), whose tenuity is bounded from below by the Buchdahl bound of $R / M = 9/4$.

@@

CAM theory, originally developed for high-energy physics applications, has been successfully applied to black hole scattering scenarios. In 1994, Andersson and Thylwe \cite{Andersson:1994rk} described the scattering of scalar waves by a Schwarzschild BH by utilising CAM techniques, and they showed that, in this case, Regge poles are associated with ``surface waves'' localised near the unstable light ring at $r=3M$ (later it was shown that the corresponding quasinormal modes are Breit-Wigner type resonances generated by these surface waves \cite{Decanini:2002ha}). Recently, Folacci and Ould Elhadj have shown that CAM methods can be used for precise numerical calculations of scattering cross sections of scalar fields, electromagnetic fields \cite{Folacci:2019cmc} and gravitational waves  \cite{Folacci:2019vtt} on the Schwarzschild spacetime. Moreover, they have obtained an approximate  formula (see Sec.~IV in Ref.~\cite{Folacci:2019vtt}) that accurately describes both the `orbiting' and `glory' oscillations in black hole scattering cross sections in the short-wavelength regime. This approximation arises from applying WKB theory to characterise the Regge pole with the smallest imaginary part, and its residue \sam{Mohamed, please correct this if wrong.}. An accurate approximation for the black hole absorption cross section has also been derived via an analysis of Regge poles \cite{Decanini:2011xi, Decanini:2011xw}.

CAM techniques are yet to be applied to compact-body scattering, it appears, with one notable exception: in 1991, Chandrasekhar and Ferrari \cite{ChandrasekharIV:1992ey} examined the Regge pole spectra for relativistic stellar models. 
Their study focussed on resonant modes (poles with small imaginary parts corresponding to trapped $w$-modes or fluid modes), and it employed CAM theory to calculate the flow of gravitational energy through a star. It seems reasonable to expect that CAM can aid our understanding of other compact objects such as and neutron stars and UCOs, and this is our purpose below.

There is a close relationship between Regge poles and quasi-normal modes (QNMs); both are sets of poles of the scattering matrix, with the latter lying in the complex-frequency plane. The excitation of black hole QNMs leads to distinctive signatures in time-dependent scattering scenarios. For example, in the immediate aftermath of a black hole merger, the perturbed black hole returns to a quiescent state via a `ringdown phase' in which its gravitational-wave signal is well described as a sum of quasinormal modes \cite{Giesler:2019uxc}. The QNM spectrum of compact bodies have been well-studied by several authors  \cite{Detweiler:1985zz, Kokkotas:1986gd, Chandrasekhar449, Kokkotas:1992ka, Leins:1993zz, Andersson:1995ez, Andersson1996}, and are reviewed in Ref.~\cite{Kokkotas:1999bd}.

It was recently shown \cite{Dolan:2017rtj, Stratton:2019deq} that time-independent scattering by compact bodies generates (in principle) a \emph{rainbow scattering} phenomenon in the short wavelength regime ($M \omega \gg 1$). The rainbow scattering phenomenon is linked to the formation of a cusp caustic in the incident wavefront (see Fig.~1 in Ref.~\cite{Stratton:2019deq}), and to the stationary point in the geodesic deflection function associated with a ray (that is, a null geodesic) that passes somewhat inside the body. For dilute bodies, such as stars and white dwarfs, the cusp caustic forms at a distance $d \sim R^2 / (4M)$ downstream of the body, and the rainbow angle is $\theta_r \approx 4M/R$ (e.g.~for the Sun, $d \approx 550 \, \text{a.u.}$ and $\theta_r \approx 1.8 \, \text{arcsec}$) \footnote{The rainbow angle $\theta_r \approx 4M/R$ is distinct from the Einstein ring angle of $\theta_E \approx \sqrt{4M/r}$}. For compact bodies such as neutron stars, however, the cusp caustic forms near the surface of the body, and the rainbow angle is large ($\theta_r \gtrsim 59.6^\circ$ for $R/M = 6$; see Table I in Ref.~\cite{Stratton:2019deq}). In this case, the rainbow angle is -- in principle at least -- a useful diagnostic of the matter distribution within the body and thus its nuclear equation of state.

@@

Finally a paragraph on the connections to Mie scattering by a transparent sphere?

\cite{Leite:2017zyb}
